NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-01 Manuscript/manuscriptPGPE.tex
Update manuscript

\clearpage
\section{Results and Discussion}

%\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}
\subsection{Conformational ensembles of different lipid headgroups in bulk bilayer}

To experimentally characterize the differences in lipid headgroup conformational ensembles
in liquid bilayer phase, we measured the C-H bond order parameters magnitudes and signs
from POPG and POPE bilayers, and compared those to our previously published results
for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters}.
%Our experimental order parameters, including the signs, from different lipid headgroups
%combined with the literature data are shown in Fig. \ref{HGorderParameters}.
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.
The most distinct order parameters are observed for PS and PG headgroups.
In PS, the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup,
%the $\alpha$-carbon order parameter is similar to PE and PG, while
the $\beta$-carbon order parameter has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
  \caption{\label{HGorderParameters}
    (top) Chemical structure of different lipids.
    (bottom) Headgroup and glycerol backbone order parameters
    from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
  }
  \todo{The bottom figure could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure}

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
The dihedral distributions show significant similarity between different lipids for all the other
bonds, except for the last two in the headgroup end, O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds,
in figure \ref{structures}. Therefore, we conclude that the disctinct order parameters of
PS and PG headgroups can be explained by the differences in these two dihedrals with respenct to other lipids.
The differences between PC and PE lipids, especially in O$_\alpha$-C$_\alpha$ bond dihedral,
may be an artefact because the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field \cite{botan15}.
In larger scale, these differences are reflected, for example, to the 
angle between headgroup dipole and membrane normal,
which decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures}).

@@

Dihedrals of both O-P bonds can freely rotate in all lipids as all possible angles
are observed in the distributions. For PS, however, some angles are quite unlikely,
possible explaining the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.
The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

@@

In conclusion, the results suggest that free rotations around O-P bonds decouple the
headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite significant differences in dihedral distributions near the headgroup end,
all lipids have dihedral angles within approximately the same ranges.
This suggests that all lipids can quite freely arrange to multiple headgroup conformations
when interaction with proteins, ions or other biomolecules.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/PCPEPGPSstructures2.png}
  \includegraphics[width=8.0cm]{../Figs/CHARMMfromLIPIDS.eps}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals1.png}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals2.png}
  \includegraphics[width=8.0cm]{../Figs/PNangleNORM.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALS.eps}
  \caption{\label{structures}
    Overlayed snapshots and dihedral angle distributions from CHARMM36 simulations of different lipids
    which give the best agreement with experiments.
  }
  \todo{More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

\clearpage
\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}
Phosphatidylcholine headgroup conformational ensemble and orientation are unchanged
upon addition of neutral molecules such as cholesterol or sphingomyelin into a bilayer,
while incorporation of charges affect the headgroup dipole vector tilt which can
be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons \cite{??}.
Such changes have been observed upon addition of charged lipids, proteins, surfactants,
drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.
%Headgroup order parameters of PC lipids decrease or increase upon addition of positive
%or negative charge into membrane, respectively, while they remain 
%\cite{??}.
%This has been explained by the tilting of the headgroup dipole due to electrostatic
%interactions with bound charges, and has been used to measure the binding affinity
%of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCvsSURFchange.eps}
  \caption{\label{changesWITHsurf}
    Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
  \caption{\label{DIHSwithSURF}
    Changes in CHARMM36 dihedrals with increasing amount of cationic surfactant.
  }
\end{figure}

The response of PC headgroup order parameters to the addition of cationic surfactants into a
membrane in CHARMM36 simulations are in very good agreement with the experimental data (Fig. \ref{changesWITHsurf})
suggesting that these simulations can be used to interpret how lipid headgroups respond to the
inclusion of charges into a membrane. Analysis of lipid dihedral bond distributions suggest that
the lipid conformational changes upon addition of charge occur in phosphate-oxygen and g$_3$-phosphate-oxygen bonds.

To resolve the lipid headgroup conformational ensembles in mixed membranes we analyzed
headgroup order parameters from mixture of PC and PE or PG and compared the changes
upon mixing to the experimental data \cite{??} (Figs. \ref{??}). However, the currently
available force fields failed to capture the headgroup conformational ensembles in mixed membranes
as we previously observed also for mixtures with PS lipids \cite{antila19}. The best performing force field for
single componen membrane, CHARMM36, overestimated the influcence of PE to PC conformations and
undersestimated response to anionic PG headgroup. The latter may arise from overbinding of sodium
counterions \cite{antila19}.

Also the changes of headgroup order parameters in POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$
%from simulations and experiments \cite{borle85,macdonald87}
in Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4} are in line with
conclusions from our previous studies \cite{catte16,melcr18,antila19,melcr20}:  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
%order parameters decrease in the response to bound cations in qualitative agreement with
%experiments, but
calcium binding affinity to membranes is typically overestimated in simulations, but
CHARMM36 with the NBfix correction underestimates the binding affinity, and that the
%On the other hand,
results can be improved by implicit inclusion of electronic polarizability.
Among the available simulations, there are two models that correctly reproduce the
PG headgroup order parameter changes upon addition of calcium, Lipid17 and Slipids,
even though the binding affinity of calcium is overestimated in these simulations.
The response of PC to calcium binding is correctly reproduced by Lipid17ecc model, as
previously reported for other membranes \cite{melcr18,melcr19}.
%to Amber based lipid models
%improves the ion binding affinity leading to a good agreement with experiments.
%(Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
%
%indicating too strong binding affinity of calcium into the bilayers as previously
%observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
%
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
%
%The calcium binding affinity to lipid bilayers with PC and PS lipids
%was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
%In this approach, the electronic polarizability is implicitly included in the classical force fields
%by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
%by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
%charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
%$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
%ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
%PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
%\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
%calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG1PC1.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (1:1) mixtures with different force fields.
  }
\end{figure}

Changes in lipid conformational ensembles characterized by distributions of heavy atom dihedral angles
are very small upon binding of calcium or addition of cationic surfactants to membranes (Figs. \ref{??}).
Yet, such changes are sufficient to tilt the headgroup dipole angle and reproduce the experimentally observed
order parameter changes in $\alpha$ and $\beta$ carbons.

\begin{figure}[]
  \centering
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPC_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
  \caption{\label{DIHSwithCA}
    Changes in POPC CHARMM36 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

>>>>

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{../Figs/figure1.eps}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of {\bf A)} POPE  and {\bf B)} POPG.
    {\bf C)} Chemical structure of POPC and POPS.
    {\bf D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
   }
   \todo{This is a sketch, needs a lot of polishing.} \\
  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

\section{Results and Discussion}

%\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}

\begin{figure*}[]
  \centering
   \includegraphics[width=18.0cm]{../Figs/figure2.eps}
%  \includegraphics[width=18.0cm]{../Figs/PCPEPGPSstructures2.png}
%  \includegraphics[width=8.0cm]{../Figs/CHARMMfromLIPIDS.eps}
%%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals1.png}
%%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals2.png}
%  \includegraphics[width=8.0cm]{../Figs/PNangleNORM.eps}
%  \includegraphics[width=8.0cm]{../Figs/DIHEDRALS.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

\subsection{Conformational ensembles of different lipid headgroups in bulk bilayer}

To experimentally characterize the differences in lipid headgroup conformational ensembles
in liquid bilayer phase, we measured the C-H bond order parameters magnitudes and signs
from POPG and POPE bilayers.
Order parameters for POPE are in good agreement with previous $^2$H NMR experiments \cite{seelig76}
and similar to our previous results for POPC \cite{ferreira16}.
In POPG, the $\beta$ and g$_2$ carbons have similar chemical environment and their peaks
overlap in the NMR spectra (Fig. \ref{HGorderParameters} B).
The signs of these order parameters were solved using SIMPSON simulations to interpre the
S-DROSS experiments in similar fashion as we did previously for PS lipids \cite{antila19}.
The resulting order parameters are compared with our previously published results
for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters} D.

%Our experimental order parameters, including the signs, from different lipid headgroups
%combined with the literature data are shown in Fig. \ref{HGorderParameters}.
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.

The most distinct order parameters are observed for PS and PG headgroups.
In PS, the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup,
%the $\alpha$-carbon order parameter is similar to PE and PG, while
the $\beta$-carbon order parameter has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.

@@

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
The dihedral distributions in figure \ref{structures} show significant similarity between different lipids for all the other
bonds, except for the last two bonds in the headgroup end, O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$.
These differences explaing the disctinct order parameters of
PS and PG headgroups with respenct to the other lipids.
The difference between PC and PE lipids in the O$_\alpha$-C$_\alpha$ bond dihedral
may be an artefact because the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field \cite{botan15}.
The differences between lipids are reflected also to the 
angle between headgroup dipole and membrane normal,
which decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures}).

@@

The analysis of dihedral angle distributions reveal the free rotation of dihedral around
phosphorus-oxygen bonds in all lipids as all possible angles are observed in the distributions.
Results in figure \ref{structure} support the recently proposed models where this free rotation 
decouples the dynamics and conformational ensembles between acyl chains and headgroup in all lipids containing
the phosphorus group \cite{antila21}. 
However, some dihedral angles of phosphorus-oxygen bonds are less likely for PS, which
%suggests that the decoupling may be weaker for PS lipids
possibly explains the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.

%The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
%are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

@@

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/figure3.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with experimental data \cite{??}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
%the results suggest that free rotations around O-P bonds decouple the
%headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
%althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite the differences in dihedral distributions of two the last bonds at the headgroup end,
leading to the distinct order parameters for PS and PG lipids,
lipids with different headgroup sample dihedral angles within approximately same ranges.
Altogether, our results suggest that lipid headgroups and very flexible thereby being
able to adopt multiple conformations when interacting with proteins, ions or other biomolecules.
Furthermore, approximately same conformations are present liquid lamellar phase for different
headgroups studied here, althought the probability weight for the conformations are slightly different.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}
%Phosphatidylcholine headgroup conformational ensemble and orientation are unchanged
%upon addition of neutral molecules such as cholesterol or sphingomyelin into a bilayer,
%while
Order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup are known to depend
linearly on the incorporation of charged molecules into a lipid bilayer,
which is explained by tilting of the headgroup dipole orientation \cite{??}.
%are kn affect the headgroup dipole vector tilt which can
%be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons.
Such changes have been observed upon addition of charged lipids, proteins, surfactants,
drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.
%Headgroup order parameters of PC lipids decrease or increase upon addition of positive
%or negative charge into membrane, respectively, while they remain 
%\cite{??}.
%This has been explained by the tilting of the headgroup dipole due to electrostatic
%interactions with bound charges, and has been used to measure the binding affinity
%of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

%\begin{figure}[]
%  \centering
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
%  \caption{\label{DIHSwithSURF}
%
%  }
%\end{figure}

The experimentally measured decrease of PC headgroup order parameters upon addition of cationic surfactants into a bilayer
are well captured by CHARMM36 simulations in figure \ref{changesWITHsurf} A,
suggesting that these simulations can be used to interpret how lipid headgroup conformational ensembles
respond to the binding of positively charged into a membrane.
Charactization of conformational ensembles using heavy atom dihedral angle distributions in figure \ref{changesWITHsurf} B
reveal that choline region is essentially unchanged upon addition of carge.
The major changes upon addition of charge are observed in dihedrals related to the phosphate oxygens,
while small change is observed also in the g$_2$-g$_3$ bond in the glycerol backbone.
This result is in line with the recently proposed model suggesting that the flexible phosphate enables the headgroup
orient according to the charge accumulated into a membrane \cite{antila21}.

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    Calsium density distributions are shown in figure \ref{CAdensPG}.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

Headgroup order parameters of PC lipids respond similarly also to the addition of charged lipids in experiments \cite{??}.
To resolve the conformational ensembles of lipid headgroups also in biologically relevant
mixtures, % of zwitterionic and charged lipids,
we compared the response of headgroup order parameters in
PC lipids to the addition of PE or PG (Fig. \ref{??}),
as well as in PG lipids to the addition of PC (Fig. \ref{??}),
between simulations and experiments. However, the accuracy of currently available force fields
is not sufficient to capture the headgroup conformational ensembles
in mixed membranes: 
The best performing force field for single componen membranes, CHARMM36, overestimates
the effect of PE to PC conformations and underestimates the response PC to the addition
of anionic PG headgroup. Previously we explained similar result for PC/PS mixtures  
%as we previously observed also for mixtures with PS lipids \cite{antila19}.
%The latter may arise from
by overbinding of the sodium counterions \cite{antila19}.

In order to analyze how conformational ensemble of PC and PG lipids response to the
bound ions, we compared the changes of headgroup order parameters
in POPC:POPG (1:1) and (4:1) mixtures upon addition of CaCl$_2$ from different simulations
to the experimental data available in the literature \cite{borle85,macdonald87}
in figures \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4}.
As in our previous studies \cite{catte16,melcr18,antila19,melcr20},  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
%order parameters decrease in the response to bound cations in qualitative agreement with
%experiments, but
the calcium binding affinity to membranes is typically overestimated in simulations,
except by CHARMM36 with the NBfix correction which underestimates the binding affinity,
and the implicit inclusion of electronic polarizability improves the results.
Lipid17ecc model with the implicit inclusion of electronic polarizability gives
the most realistic response of PC lipid headgroup order parameters to the binding of calcium.
In this model, the main effect of calcium to the lipid conformational ensemble is
%the decrease is headgroup order parameters observed in this model is
the slight change of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral distribution from trans state
to eclipsed conformations (Fig. \ref{DIHSwithCAlipid17eccPOPC}).
This is in line with the changes observed in CHARMM36 simulations upon addition of charged surgactants (Fig. \ref{changesWITHsurf}),
despite the major differences in lipid headgroup conformational ensembles between
these models without ions (Fig. \ref{structures} vs. Fig. \ref{DIHSwithCAlipid17eccPOPC}).
%of lipid headgroup is not 
%Changes in lipid conformational ensembles characterized by distributions of heavy atom dihedral angles
%are very small upon binding of calcium or addition of cationic surfactants to membranes (Figs. \ref{??}).
%Yet, such changes are sufficient to tilt the headgroup dipole angle and reproduce the experimentally observed
%order parameter changes in $\alpha$ and $\beta$ carbons.

%Among the available simulations, there are two models that correctly reproduce the
%PG headgroup order parameter changes upon addition of calcium, Lipid17 and Slipids,
%even though the binding affinity of calcium is overestimated in these simulations.
%The response of PC to calcium binding is correctly reproduced by Lipid17ecc model, as
%previously reported for other membranes \cite{melcr18,melcr19}.

%to Amber based lipid models
%improves the ion binding affinity leading to a good agreement with experiments.
%(Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
%
%indicating too strong binding affinity of calcium into the bilayers as previously
%observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
%
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
%
%The calcium binding affinity to lipid bilayers with PC and PS lipids
%was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
%In this approach, the electronic polarizability is implicitly included in the classical force fields
%by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
%by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
%charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
%$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
%ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
%PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
%\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
%calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.
